---
title: "EDA of Four Horses in a Race"
output: html_document
author: "150000760"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Executive Summary

Our report finds that the maximum amount of winnings possible only betting before the race on the winning horse is 11976.60 GBP, taking a lay position that would cover the back stake, if the winning horse had actually lost. The total lay and back stakes were 70.28 + 1303.47 = 1373.75, so a total percentage profit of 770%. 

The expectation and variance shows that gamblers price things somewhat correctly, that is they give each horse an equal shot, and have tremendous anxiety. 

Relevant plots are redone in SAS, to prove repeatability. 

### Introduction

This report analyzes betting exchange data for three randomly chosen horses and the winner. After data cleaning and Exploratory Data Analysis we determine the maximum amount of profit that could be made using an arbitrage position for the winner and the losers. 

Backing a horse is betting that that horse will win. Laying a horse is betting that that horse will lose. Given n horses in a race, n-1 of them will lose, so the probability that a horse loses is higher than the probability that a given horse wins. A stake is the amount of money you will place on a bet, which is capped by the total volume in the market. 

Odds are the multiple offered on a stake for a given bet, laying or backing. Odds given are intrinsically related to the probability that a certain event will occur. This is because the goal of a fair bet is to make the expected value zero. 

A beautiful analysis on whether such gambling is worth doing is Daniel Bernoulli's "Exposition of a New Theory on the Measurement of Risk" from 1738. It is available on JSTOR. 

## Exploratory Data Analysis

```{r echo = FALSE}
library(readxl)
HorseData <- read_excel("~/Desktop/MT5763/Project1/HorseData.xlsx")
```


With any dataset, it is good to see summary stats. 

```{r echo = FALSE}
summary(HorseData)
```

The data is not nice! We would like Price and Volume columns to be numeric so that we may manipulate such calculations. 

```{r echo = FALSE}
HorseData$backPrice1 <- as.numeric(HorseData$backPrice1)
HorseData$backVolume1 <- as.numeric(HorseData$backVolume1)
HorseData$layPrice1 <- as.numeric(HorseData$layPrice1)
HorseData$layVolume1 <- as.numeric(HorseData$layVolume1)
```


There is quite a number of NaNs! Can we visualize them? 

```{r}
library(naniar)
vis_miss(HorseData, warn_large_data = FALSE)
```

For Ease of analysis, we would like to remove NaN values, but important information - whether the Horse won or lost - is in these columns. A Column is formed of all False Booleans, then the competitor 11473056 with the winning horse is changed to true. 

```{r}
HorseData$Winner <- FALSE
HorseData$Winner[HorseData$competitor == 11473056] <- TRUE

```
Now the NaNs are removed. 

```{r}
HorseData <- na.omit(HorseData)
```

And finally, two new columns are added for ease of analysis
```{r}
HorseData$backVP <- HorseData$backPrice1 * HorseData$backVolume1
HorseData$layVP <- HorseData$layPrice1 * HorseData$layVolume1
```

```{r}
summary(HorseData)
```


## Variance and Expectation analysis

There are in total, 33 unique competitors in the two races. 

```{r}
unique(HorseData$competitor)
```


```{r}
sorted_data <- HorseData[order(HorseData$competitor, HorseData$time),]
```

The following four competitors were randomly chosen. For ease of data replication, the extracted datasets were exported to CSV. 
```{r output=FALSE}
winner<-sorted_data[which(sorted_data$competitor==11473056), ]
write.csv(winner,"./winner.csv", row.names=FALSE)
competitor2<-sorted_data[which(sorted_data$competitor==11538828), ]
write.csv(competitor2,"./competitor2.csv", row.names=FALSE)
competitor3<-sorted_data[which(sorted_data$competitor==9977366), ]
write.csv(competitor3,"./competitor3.csv", row.names=FALSE)
competitor4<-sorted_data[which(sorted_data$competitor==10339376), ]
write.csv(competitor4,"./competitor4.csv", row.names=FALSE)
```

Back and Lay odds trace for pre-inplay data. 

```{r echo = FALSE}

winneroutplay<-sorted_data[which(winner$inplay==FALSE,), ]
plot(winneroutplay$time,winneroutplay$backPrice1)
competitor2outplay<-competitor2[which(competitor2$inplay==FALSE,), ]
plot(competitor2outplay$time,competitor2outplay$backPrice1)
competitor3outplay<-competitor3[which(competitor3$inplay==FALSE,), ]
plot(competitor3outplay$time,competitor3outplay$backPrice1)
competitor4outplay<-competitor4[which(competitor4$inplay==FALSE,), ]
plot(competitor4outplay$time,competitor4outplay$backPrice1)
```


```{r}
plot(winneroutplay$time,winneroutplay$layPrice1)
plot(competitor2outplay$time,competitor2outplay$layPrice1)
plot(competitor3outplay$time,competitor3outplay$layPrice1)
plot(competitor4outplay$time,competitor4outplay$layPrice1)
```
### Mean/Variance for each player without normalisation. 


Winner mean and variance 
```{r}
mean(winneroutplay$layPrice1)
var(winneroutplay$layPrice1)
mean(winneroutplay$backPrice1)
var(winneroutplay$backPrice1)
```
Competitor 2 mean and variance

```{r}
mean(competitor2outplay$layPrice1)
var(competitor2outplay$layPrice1)
mean(competitor2outplay$backPrice1)
var(competitor2outplay$backPrice1)
```
Competitor 3 mean and variance

```{r}
mean(competitor3outplay$layPrice1)
var(competitor3outplay$layPrice1)
mean(competitor3outplay$backPrice1)
var(competitor3outplay$backPrice1)
```
Competitor 4 mean and variance
```{r}
mean(competitor4outplay$layPrice1)
var(competitor4outplay$layPrice1)
mean(competitor4outplay$backPrice1)
var(competitor4outplay$backPrice1)
```


Mean and variance do not tell you much. However competitor 3 and 4 appear to be more volatile, and reflects uncertainity in the markets. 


## How much money could you have made on the winner?

```{r echo = FALSE}
plot(winneroutplay$time,winneroutplay$backVP)
```


```{r echo = FALSE}

max(winneroutplay$backVP)
winneroutplay$time[which.max(winneroutplay$backVP)]
winneroutplay$backVolume1[which.max(winneroutplay$backVP)]
```
So the most money you could have made is by backing the winner at 15:56:50.

What was the stake you needed?

```{r echo = FALSE}
winneroutplay$backVolume1[which.max(winneroutplay$backVP)]
```

So could you have found a lay, so that the 437.14 stake would have been recovered if you lost?

```{r echo = FALSE}
winneroutplay$time[which.min(winneroutplay$layVP>1303.47)]
winneroutplay$layVolume1[which.min(winneroutplay$layVP>1303.47)]
winneroutplay$layPrice1[which.min(winneroutplay$layVP>1303.47)]
```
So your overall earnings if you backed and laid at the time indicated above would be 11976.2 

```{r echo = FALSE}
max(winner$backVP)-winner$layVP[which.min(winner$layVP>1303.47)]
```
Your overall return would be 11976.2 on your initial investment. Not so bad!



