---
title: "EDA of Four Horses in a Race"
author: '150000760'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Executive Summary

Our report finds that the maximum amount of winnings possible only betting before the race on the winning horse is 12,698.87. Your initial investment was 437.14+22.45 = 459.59, so your percentage return is 2750%.

The expectation and variance reflects uncertainty to predictions, which is right as gamblers should be uncertain. 

Relevant plots are redone in SAS, to prove repeatability, and are attached in a seperate file . 

### Introduction

This report analyzes betting exchange data for three randomly chosen horses and the winner. After data cleaning and Exploratory Data Analysis we determine the maximum amount of profit that could be made using an arbitrage position for the winner and the losers. 

Backing a horse is betting that that horse will win. Laying a horse is betting that that horse will lose. Given n horses in a race, n-1 of them will lose, so the probability that a horse loses is higher than the probability that a given horse wins. A stake is the amount of money you will place on a bet, which is capped by the total volume in the market. 

Odds are the multiple offered on a stake for a given bet, laying or backing. Odds given are intrinsically related to the probability that a certain event will occur. This is because the goal of a fair bet is to make the expected value zero. 


## Exploratory Data Analysis

```{r echo = FALSE}
library(readxl)
HorseData <- read_excel("~/Desktop/MT5763/Project1/HorseData.xlsx")
```


With any dataset, it is good to see summary stats. 

```{r echo = FALSE}
summary(HorseData)
```

The data is not nice! We would like Price and Volume columns to be numeric so that we may manipulate such calculations. 

```{r echo = FALSE}
HorseData$backPrice1 <- as.numeric(HorseData$backPrice1)
HorseData$backVolume1 <- as.numeric(HorseData$backVolume1)
HorseData$layPrice1 <- as.numeric(HorseData$layPrice1)
HorseData$layVolume1 <- as.numeric(HorseData$layVolume1)
```


There is quite a number of NaNs! Can we visualize them? 

```{r}
library(naniar)
vis_miss(HorseData, warn_large_data = FALSE)
```

For Ease of analysis, we would like to remove NaN values, but important information - whether the Horse won or lost - is in these columns. A Column is formed of all False Booleans, then the competitor 11473056 with the winning horse is changed to true. It does not matter if you remove the final backPrice1 rows. By that point the horse would have lost anyways, so it would not be possible to lay them.

```{r }
HorseData$Winner <- FALSE
HorseData$Winner[HorseData$competitor == 11473056] <- TRUE

```
Now the NaNs are removed. 

```{r}
HorseData <- na.omit(HorseData)
```

And finally, two new columns are added for ease of analysis
```{r}
HorseData$backVP <- HorseData$backPrice1 * HorseData$backVolume1
HorseData$layVP <- HorseData$layPrice1 * HorseData$layVolume1
```

```{r}
summary(HorseData)
```


## Variance and Expectation analysis

There are in total, 33 unique competitors in the two races. 

```{r}
unique(HorseData$competitor)
```


```{r}
sorted_data <- HorseData[order(HorseData$competitor, HorseData$time),]
```

The following three competitors and the winner were randomly chosen (by eyeball not RNG). For ease of data replication, the extracted datasets were exported to CSV. 
```{r output=FALSE}
winner<-sorted_data[which(sorted_data$competitor==11473056), ]
write.csv(winner,"./winner.csv", row.names=FALSE)
competitor2<-sorted_data[which(sorted_data$competitor==11538828), ]
write.csv(competitor2,"./competitor2.csv", row.names=FALSE)
competitor3<-sorted_data[which(sorted_data$competitor==9977366), ]
write.csv(competitor3,"./competitor3.csv", row.names=FALSE)
competitor4<-sorted_data[which(sorted_data$competitor==10339376), ]
write.csv(competitor4,"./competitor4.csv", row.names=FALSE)
```

# Back and Lay odds trace for pre-inplay data. 

```{r echo = FALSE}

winneroutplay<-winner[which(winner$inplay==FALSE & winner$marketStatus=="OPEN"), ]
write.csv(winneroutplay,"./winneroutplay.csv", row.names=FALSE)
plot(winneroutplay$time,winneroutplay$backPrice1,
     main="Winner Back Price over Time",
     xlab = "Time",
     ylab = "Back Price")

competitor2outplay<-competitor2[which(competitor2$inplay==FALSE & competitor2$marketStatus=="OPEN"), ]


write.csv(competitor2outplay,"./competitor2outplay.csv", row.names=FALSE)
plot(competitor2outplay$time,competitor2outplay$backPrice1,
     main="Competitor 2 Back Price over Time",
     xlab = "Time",
     ylab = "Back Price")
competitor3outplay<-competitor3[which(competitor3$inplay==FALSE& competitor3$marketStatus=="OPEN"), ]

write.csv(competitor3outplay,"./competitor3outplay.csv", row.names=FALSE)
plot(competitor3outplay$time,competitor3outplay$backPrice1,
     main="Competitor 3 Back Price over Time",
     xlab = "Time",
     ylab = "Back Price")
competitor4outplay<-competitor4[which(competitor4$inplay==FALSE & competitor4$marketStatus=="OPEN"), ]

write.csv(competitor4outplay,"./competitor4outplay.csv", row.names=FALSE)
plot(competitor4outplay$time,competitor4outplay$backPrice1,
     main="Competitor 4 Back Price over Time",
     xlab = "Time",
     ylab = "Back Price")


```


```{r}
plot(winneroutplay$time,winneroutplay$layPrice1,
     main="Winner Lay Price over Time",
     xlab = "Time",
     ylab = "Lay Price")
plot(competitor2outplay$time,competitor2outplay$layPrice1,
     main="Competitor 2 Lay Price over Time",
     xlab = "Time",
     ylab = "Lay Price")
plot(competitor3outplay$time,competitor3outplay$layPrice1,
     main="Competitor 3 Lay Price over Time",
     xlab = "Time",
     ylab = "Lay Price")
plot(competitor4outplay$time,competitor4outplay$layPrice1,
     main="Competitor 4 Lay Price over Time",
     xlab = "Time",
     ylab = "Lay Price")
```


## Mean/Variance for each player without normalisation. 


### Winners mean and variance 
```{r echo = FALSE}
print("Lay Price Mean")
mean(winneroutplay$layPrice1)
print("Lay Price Variance")
var(winneroutplay$layPrice1)
print("Back Price Mean")
mean(winneroutplay$backPrice1)
print("Back Price Variance")
var(winneroutplay$backPrice1)
```
### Competitor 2 mean and variance

```{r echo = FALSE}
print("Lay Price Mean")
mean(competitor2outplay$layPrice1)
print("Lay Price Variance")
var(competitor2outplay$layPrice1)
print("Back Price Mean")
mean(competitor2outplay$backPrice1)
print("Back Price Variance")
var(competitor2outplay$backPrice1)
```
### Competitor 3 mean and variance

```{r echo = FALSE}
print("Lay Price Mean")
mean(competitor3outplay$layPrice1)
print("Lay Price Variance")
var(competitor3outplay$layPrice1)
print("Back Price Mean")
mean(competitor3outplay$backPrice1)
print("Back Price Variance")
var(competitor3outplay$backPrice1)
```
### Competitor 4 mean and variance
```{r echo = FALSE}
print("Lay Price Mean")
mean(competitor4outplay$layPrice1)
print("Lay Price Variance")
var(competitor4outplay$layPrice1)
print("Back Price Mean")
mean(competitor4outplay$backPrice1)
print("Back Price Variance")
var(competitor4outplay$backPrice1)
```


Mean and variance without standardization do not tell you much. To an extent, it reflects uncertainty and how the markets guage certain horses. By this metric, it seems competitor 3 and 4 were favorites, though with large uncertainty, wheras the winner was not a favorite, though the markets did not rule him out. Visually, there is increased movement right before the race starts. Without using statistical tests, one sees that it is more "jagged" and "movey" then. 


## How much money could you have made on the winner?

Here is a plot of volume*price. The maxima reflects the amount of money you could have made if you bought at that specific time, the entire volume.

```{r echo = FALSE}
plot(winneroutplay$time,winneroutplay$backVP)
```

Lets find the maximum vp, the time at which it occured, and the maximum stake size. 

```{r echo = FALSE}

max(winneroutplay$backVP)
winneroutplay$time[which.max(winneroutplay$backVP)]
winneroutplay$backVolume1[which.max(winneroutplay$backVP)]
winneroutplay$backPrice1[which.max(winneroutplay$backVP)]
```
So 13,114.20 GBP was the maximum you could have made if you backed the winner at 15:56:50 with a 437.14 stake at 30/1 odds.


So could you have found a lay, so that the 437.14 stake would have been recovered if you lost?

```{r echo = FALSE}
# find min value
n = min(winneroutplay$layVP[which(winneroutplay$layVP>437.14)])
n
# find index of min value

m = which(winneroutplay$layVP==n)[1]

winneroutplay$time[m]
winneroutplay$layVolume1[m]
winneroutplay$layPrice1[m]
```
If you laid the horse at "2018-06-20 15:41:50 UTC" with a 22.45 GBP stake at 19.5 odds, you would make 437.78 GBP and covered your loss. You are however liable for the remaining 437.78-22.45 = 415.33 GBP if the horse wins. 

So the horse won and you got 13114.20. You have to pay 415.33 out of your winnings, which leaves you at 12,698.87. Your initial investment was 437.14+22.45 = 459.59, (or 874.92 if you include your liability), so your percentage return is 2750%. 


## Bibliography

Bernoulli Article on Risk 
https://www.jstor.org/stable/1909829?seq=1#metadata_info_tab_contents

Explained Betting Exchange
https://betting.betfair.com/what-is-a-betting-exchange.html

Explains how laying work
https://betting.betfair.com/how-to-use-betfair-exchange/beginner-guides/placing-a-lay-bet-010819-51.html

